---
title: "6. Exporting data"
author:
  affiliation: Centre for Research into Ecological & Environmental Modelling, University of St Andrews
  name: Phil Bouchet, Enrico Pirotta, Len Thomas, Catriona Harris
date: "`r Sys.Date()`"
css: pkgdown/extra.css
csl: narwind.csl
link-citations: yes
rmarkdown::html_vignette:
  fig_caption: yes
  toc: yes
  toc_depth: 4
bibliography: narwind.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{6. Exporting data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

```{r, echo = FALSE}
downloadthis::download_link(
  link = "https://raw.githubusercontent.com/offshore-wind/narwind/0a45ae03f1938cd5396895b943b8615f28812669/vignettes/pdf/6%20Exporting%20data%20%20narwind.pdf",
  button_label = "Download PDF",
  button_type = "primary",
  has_icon = TRUE,
  icon = "fa fa-file-pdf",
  self_contained = FALSE
)
```

### Exporting data {data-link="Exporting data"}

There are two ways to save the results of an analysis conducted in `narwind`.

The first is to save the R objects returned by the model as an `.rda` (or 'R data') file using the `save()` command. **This is essential** as running the agent-based model can be time-consuming and repeat analyses should therefore be avoided. To do this, execute the following line of code, making sure to replace the path and fine name with appropriate values:

```{r, eval = FALSE}
save(model_base, file = "path/to/a/folder/of/choice/filename.rda")
```

During package testing, a typical `.rda` file containing outputs from a run of the agent-based model with `nsim = 1000` was between 500 Mb and 1 Gb in size.

In addition to the above, the `export()` function can be used to save model outputs as a Microsoft Excel spreadsheet (.xlsx), stored in the user's current working directory. `export()` can be used with both outputs from the agent-based model and the population model. In the former case, a separate file will be produced for each cohort in `cohort`, so we recommend using this option only for short simulation runs (`nsim = 100`), or using the `cohort` and `whale` arguments to extract data for specific individuals/cohorts.

Note that if the simulation run was given a label when using `narw()`, then this label is appended to the output file name.

In the below example, the first line of code exports data from the agent-based model for the first 10 females in the lactating cohort (ID = 5). The second line of code exports population projection data.

```{r save, eval=FALSE}
export(model_base, whale = 1:10, cohort = 5)
export(proj_base, prefix = "NARW_abundance")
```

All arguments to `export()` are listed below.

| Object class | Argument | Default value | Description                                                                                                                 |
|----------|----------|----------|--------------------------------------------|
| `narwsim`    | `obj`    | `-`           | Input model object, as returned by `narw()`.                                                                                |
|              | `prefix` | `"narwsim"`   | Character string. Prefix appended to the output file name.                                                                  |
|              | `whale`  | `1:nsim`      | Positive integer indicating the individual(s) for which data should be extracted. By default, all individuals are included. |
|              | `cohort` | `1:6`         | Positive integer or vector of positive integers indicating which cohort(s) should be considered. Defaults to all cohorts.   |
| `narwproj`   | `obj`    | `-`           | Input projection object, as returned by `predict()`.                                                                        |
|              | `prefix` | `"narwproj"`  | Character string. Prefix appended to the output file name.                                                                  |
|              | `...`    |               | Additional arguments passed to `write.xlsx`                                                                                 |
